Scaling Automated Theory Exploration

### 2015-08-06 AI4FM/article.tex
Added diagram

By removing user interaction, automated TE systems require an algorithm
for deciding whether to inform the user of a particular theorem or not; we refer
to this search criterion as the theorem being ``interesting''.

In existing systems, this criterion is intimately connected to the
search algorithm; this coupling improves efficiency, even for brute-force
search. As an example, \textsc{IsaCoSy} \cite{johansson2009isacosy} discovers
equations, which are defined as ``interesting'' if they cannot be simplified
using previously discovered equations. The intuition for such criteria is to
avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$,
etc. when we already know $\forall x. 0 + x = x$. To work effectively, this
requires general forms to be found \emph{before} special cases.

In the following, we give an overview of the state of the art in automated
theory exploration; identify potential improvements; and present our initial
attempts to implement some of these ideas.

\section{Theory Exploration in Haskell}

Automated theory exploration has been applied to libraries in Isabelle
and Haskell, although we focus on the latter as its implementations are
the most mature (demonstrated by the fact that \textsc{Hipster} explores
Isabelle by first translating it to Haskell). Haskell is interesting to target,
since its use of pure functions and algebraic datatypes causes many functions to
follow algebraic laws; however, since Haskell's type system cannot encode such
laws (without difficulty \cite{lindley2014hasochism}, at least), less
effort is given to finding and stating them; compared to a theorem proving
setting like Isabelle, where these laws are often the main focus.

Due to Haskell's relative popularity, there are also large code repositories
such as \textsc{Hackage} available to explore. This gives several potential
benefits: in the short term, existing library authors and users may be assisted
in comprehending and maintaining their code \cite{QuickSpec}; more notably, TE
could be used to data mine new abstractions, such as typeclasses, with relevance
to real-world code.

Currently, the most powerful TE system for Haskell is \textsc{HipSpec}.
This uses \textsc{QuickSpec} to search through \emph{expressions}
(combinations of the Haskell terms given by the theory), rather than
searching through the space of equations or proofs directly. Expressions
are grouped into equivalence classes, such that the \textsc{QuickCheck}
counterexample finder cannot distinguish between the elements; equations
relating the members of these classes are then conjectured, and sent to
existing automated theorem provers to try and prove \cite{rosen2012proving}. This
approach works well as a lemma generation system, making
\textsc{HipSpec} a capable inductive theorem prover as well as a theory
exploration system \cite{claessen2013automating}.

\section{Identifying Improvements}

Even with a slow search algorithm, we can use a divide and conquor
approach to limit the number of allowed combinations, either by using
stricter types to prevent composition, or by partitioning the theory
into small independently-searched sub-theories. Of course, such
restrictions should strike a balance between the efficiency gained and
the potential to forbid some interesting theorems.

Given this state of the art, we identify the following as potential
areas for improvement:

For a more thorough treatment,
surveys recent work on search algorithms and  surveys
interestingness criteria in the context of data mining.

\begin{itemize}
\item
  Enumerating expressions is a brute-force approach. Identifying more scalable
  alternatives to brute-force search is a well-studied area of Artificial
  Intelligence and Machine Learning. In particular, there is potential in
  applying heuristic algorithms like those surveyed in \cite{blum2011hybrid}.
\item
  Many ``interestingness'' criteria are surveyed in
  \cite{geng2006interestingness}, but existing systems are tightly coupled to
  one particular notion. Whilst pre- and post-processors could provide
  additional filtering, there is merit in building a more loosely-coupled
  alternative, for two reasons:

  \begin{itemize}
  \item
    As the number of irreducible equations grows, it may be desirable to
    impose extra conditions of a more subjective nature.
  \item
    Surprising, insightful equations may be discarded by current systems, if
    they are actually reducible in some complex, non-obvious way. A more
    subjective interestingness measure could be used to veto such
    rejections.
  \end{itemize}
\item
  HipSpec does not propose candidate equations by data mining
  previous results; generalisation methods like anti-unification could
  do this, and at the same time remove the requirement that general
  forms must be enumerated early.
\item
  All type-safe combinations of the given expressions are tried, whilst
  it may be discernable a priori that some combinations are not worth
  considering (either because they are never related, or because their
  relations are never interesting).
\end{itemize}

@@

\section{The \textsc{ML4HS} Framework}

We are implementing a system called \textsc{ML4HS} to investigate scalable
automated theory exploration for Haskell. At this stage, we provide a
pre-processor for \textsc{QuickSpec} which attempts to make exploration of large
theories more tractable by splitting them into a collection of smaller theories.

Inspired by premise selection in automated theorem proving
\cite{kuhlwein2012overview}, and similarity-based clustering method such as that
of \textsc{ML4PG} \cite{journals/corr/abs-1302-6421}, we are investigating the
hypothesis that similar expressions are more likely to be related by equational
properties than dissimilar expressions. This has lead to a divide and conquor
approach, shown in figure 1.

Since our aim is to scale up theory exploration, our pre-processor also
increases the level of automation compared to existing systems. We consider
\textsc{Hackage} packages to be theories, and treat package management,
downloading, compiling, dependency resolution, etc. as necessary tasks for a TE
system such as \textsc{ML4HS} to implement.

Not only does this eliminate the need to define theories manually, as a sub-set
of available expressions, but may also be useful in its own right as a mechanism
to execute arbitrary Haskell code from arbitrary modules in arbitrary packages.

% Acknowledgements for colleagues, referees, ...
% ==============================================
\begin{acknowledge}
I am grateful for those who have helped formulate these ideas through
conversation, especially the HipSpec team at Chalmers University (Moa
Johansson, Koen Claessen, Nick Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena)
and my supervisor Katya Komendantskaya. I also wish to thank the implementors
of the systems we are building on, including HipSpec, QuickSpec and
QuickCheck on the theory exploration side, as well as GHC, Cabal and Nix
on the infrastructure side.
\end{acknowledge}

>>>>

By removing user interaction, automated TE systems require an algorithm
for deciding whether to inform the user of a particular theorem or not; we refer
to this search criterion as the theorem being ``interesting''. In existing
systems, this criterion is intimately connected to the
search algorithm; such coupling improves efficiency, even for brute-force
search, but makes it difficult to try alternative measures.

As an example, \textsc{IsaCoSy} \cite{johansson2009isacosy} discovers
equations, which are defined as ``interesting'' if they cannot be simplified
using previously discovered equations. The intuition for such criteria is to
avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$,
etc. when we already know $\forall x. 0 + x = x$. Whilst elegantly implemented
as a term rewriting system, to work effectively this requires general forms to
be found \emph{before} special cases.

In the following, we give an overview of the state of the art in automated
theory exploration (\autoref{haskell}). We identify potential improvements to
these systems (\autoref{improve}) and present our initial implementation
attempts in the form of the \textsc{ML4HS} system (\autoref{ml4hs}).

\section{Theory Exploration in Haskell}\label{haskell}

Automated theory exploration has been applied to libraries in Isabelle
and Haskell, although we focus on the latter as its implementations are
the most mature (demonstrated by the fact that \textsc{Hipster} explores
Isabelle by first translating it to Haskell). Haskell is interesting to target,
since its use of pure functions and algebraic datatypes causes many functions to
follow algebraic laws. However, since Haskell's type system cannot encode such
laws (without difficulty \cite{lindley2014hasochism}, at least); less
effort is given to finding and stating them, compared to a theorem proving
setting like Isabelle. Hence, Haskell provides much low-hanging fruit for a
broad but shallow exploration to discover.

Due to Haskell's relative popularity, there are also large code repositories
such as \textsc{Hackage} available to explore. This gives several potential
benefits: in the short term, applying TE to such repositories may assist
existing library authors and users in comprehending and maintaining their code
\cite{QuickSpec}; more notably, TE could be used to data mine new abstractions,
such as typeclasses, with relevance to real-world applications.

Currently, the most powerful TE system for Haskell is \textsc{HipSpec}.
This uses the earlier \textsc{QuickSpec} system to search through
\emph{expressions} (combinations of the Haskell terms given by the theory),
rather than searching through the space of equations or proofs directly.
Expressions are grouped into equivalence classes, such that the
\textsc{QuickCheck} counterexample finder cannot distinguish between the
elements; equations relating the members of these classes are then conjectured,
and sent to existing automated theorem provers to try and prove
\cite{rosen2012proving}. This approach works well as a lemma generation system,
making \textsc{HipSpec} a capable inductive theorem prover as well as a theory
exploration system \cite{claessen2013automating}.

\section{Identifying Improvements}\label{improve}

Given this state of the art, we identify the following as potential
areas for investigation:

\begin{itemize}
\item
  More scalable alternatives to brute-force enumeration. This is a well-studied
  area of Artificial Intelligence and Machine Learning. In particular, there is
  potential in applying heuristic algorithms like those surveyed in
  \cite{blum2011hybrid}.
\item
  Using alternative ``interestingness'' criteria, like those surveyed in
  \cite{geng2006interestingness}. In particular, we envision two problems with
  large-scale exploration which may require more sophisticated criteria:

  \begin{itemize}
  \item
    As the number of irreducible equations grows, it may be desirable to
    impose extra conditions of a more subjective nature; e.g. to distinguish
    useful relationships from incidental coincidences.
  \item
    Surprising, insightful equations may be discarded by current systems, if
    they are actually reducible in some complex, non-obvious way. A more
    subjective interestingness measure could be used to veto such
    rejections.
  \end{itemize}
\item
  Data mining previous results; for example, using generalisation methods like
  anti-unification \cite{bulychev2010anti}. This would make more efficient use
  of data, and also remove the requirement that general forms be enumerated
  early.
\item
  Divide and conquer approaches. Rather than try all type-safe combinations of
  the given expressions, it may be discernable a priori that some combinations
  are not worth considering (either because they are never related, or because
  their relations are never interesting). In other words, automating the theory
  selection process currently performed by the user.
\end{itemize}

@@

\section{The \textsc{ML4HS} Framework}\label{ml4hs}

\begin{figure}
  \centering
  \includegraphics[]{ml4hs}
  \caption{An overview of the \textsc{ML4HS} system}
  \label{ml4hsdia}
\end{figure}

We are implementing a system called \textsc{ML4HS} to investigate scalable
automated theory exploration for Haskell. The current form of \textsc{ML4HS} is
a pre-processor for \textsc{QuickSpec} which attempts to make exploration of
large theories more tractable by splitting them into a collection of smaller
theories: the divide and conquer approach mentioned above. This is
similar to reducing the search space in automated theorem proving via premise
selection \cite{kuhlwein2012overview}. An overview of \textsc{ML4HS} is given in
\autoref{ml4hsdia}. We have integrated our pre-processor with the \textsc{Cabal} and
\textsc{Nix} package managers so that \textsc{ML4HS} can use regular Haskell
packages as theories.

When selecting the components of each sub-theory, we want to keep together those
expressions which combine to form interesting theorems, and separate those which
combine in uninteresting ways. Our hypothesis is that similarity-based
clustering of expressions, as found in \textsc{ML4PG}
\cite{journals/corr/abs-1302-6421}, is an effective method for performing this
division. Future experiments will test this by measuring throughput rates of
\textsc{ML4HS} compared to \textsc{QuickSpec} and \textsc{HipSpec}.

% Acknowledgements for colleagues, referees, ...
% ==============================================
\begin{acknowledge}
Thank you to my supervisor Katya Komendantskaya and members of the
\textsc{HipSpec} team at Chalmers University (Moa Johansson, Koen Claessen, Nick
Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena) for useful discussions of these
ideas.
\end{acknowledge}
